### 2015-08-06 whoop.tex
Complete pass.

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged to an automated theorem prover.

@@

\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers, or dynamic memory allocation.  These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

@@

The allowed statements are shown in Fig.~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

@@

\noindent\textbf{Semantics }
%
Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.  The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.  Each thread consists of an id (drawn from $\ids$), a thread template $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.  If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.

Initially, the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
%
We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is held by some thread.  Executing a statement updates the thread-private and shared state in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

A thread terminates if it reaches the end of $\mainforthread{T}$, and is removed from the list of threads.  Since our interest is in analysis of device drivers, which are reactive, we do not consider the notion of global program termination.

\noindent\textbf{Lockset Instrumentation }
%
For templates $T$ and $U$, which can be equal, we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrary many further concurrently executing threads.
%
To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{sec:background}).  Given an arbitrary symbol $i$, we define the instrumentation of $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.
%
Renaming is straightforward: all occurrences of each private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $T$ and $U$ are the same.

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Fig.~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Fig.~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Fig.~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

\noindent\textbf{Architecture }
%
Fig.~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Fig.~\ref{fig:whoop}--C and \S\ref{corral}).

@@

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. Thus, \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts.

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant. Representing all locks statically, instead of their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
%
Currently, \whoop supports Linux kernel mutex and spinlock operations; it is easy to enhance it with knowledge of other locking primitives.
%
If \whoop cannot reliably infer a unique lock instance (e.g.\ because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning. However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Fig.~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

\noindent\textbf{Generating Loop and Procedure Summaries }
%
Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.  We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined.

@@

In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.

@@

\noindent\textbf{Limitations }
%
As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lock-free synchronization is used). \whoop also uses over-approximation, which can be another source of imprecision. Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the later could be addressed by providing a mechanism for users to declare custom locks. We also do not currently treat interrupt handlers in a special way; we just assume that they can execute concurrently with any entry point. One way to address this is to model interrupt-specific kernel functions (e.g. for enabling/disabling interrupts).

>>>>

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented our analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
\COULDCUT{Our experimental evaluation (\S\ref{evaluation}) demonstrates that \whoop has value as a stand-alone analyzer, and that results from our analysis can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.}

@@

Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence.  A post-condition for the sequential program asserts, for every shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions (VCs)~\cite{barnett2005weakest} are generated and discharged to an automated theorem prover.

@@

\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point.  Multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers, or dynamic memory allocation.  These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

@@

The allowed statements are shown in Fig.~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below and is trivial to enforce by pre-processing.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

@@

\noindent\textbf{Semantics }
%
Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.  The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{thread state}.  Each thread state has an associated thread template $T$, the type of the thread (and multiple threads may have the same associated template), and consistes of an id (drawn from $\ids$), an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables $\variables{T}$.  If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.

Initially, the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of thread states is empty.
%
At any point of execution, a new thread state may be added to the list of thread states.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread state to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
%
We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is held by some thread.  Executing a statement updates the thread-private and shared state in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

A thread terminates if it reaches the end of $\mainforthread{T}$, in which case its state is removed from the list of thread states.  Since our interest is in analysis of device drivers, which are reactive, we do not consider the notion of global program termination.

\noindent\textbf{Lockset Instrumentation }
%
For templates $T$ and $U$, which can be equal, we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrary many further concurrently executing threads.
%
To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{sec:background}).  Given an arbitrary symbol $i$, we define the instrumentation of $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.
%
Renaming is straightforward: each occurrence of a private variable $x \in \variables{T}$ used in $T$ is replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $T$ and $U$ are the same.

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Fig.~\ref{fig:sequentialization} is correct (i.e.\ the assertion, described above, holds invariantly at the end of the program), then it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Fig.~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program is correct, and that the threads terminate, the assertion checked at the end of the program guarantees that at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Fig.~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

@@

\noindent\textbf{Architecture }
%
Fig.~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}.  Boogie is a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ the Boogie verifier and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Fig.~\ref{fig:whoop}--B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Fig.~\ref{fig:whoop}--C and \S\ref{corral}).

@@

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation and, hence, we do not need to keep track of them separately.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. Thus, \whoop does not need to directly deal with pointers and alias analysis\COULDCUT{, a hard problem in its own right, which allows us to reuse robust existing techniques and focus instead on verification efforts}.

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant. Representing all locks statically, instead of using their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
%
Currently, \whoop supports Linux kernel mutex and spinlock operations; it is easy to enhance it with knowledge of other locking primitives.
%
If \whoop cannot reliably infer a unique lock instance (e.g.\ because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning. However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held when accessing each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Fig.~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A\;.\; \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

\noindent\textbf{Generating Loop and Procedure Summaries }
%
Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.  We found that this did not scale to large drivers, and some drivers do exhibit recursion, e.g.\ the r8169 ethernet driver (see \S\ref{evaluation}).

@@

\COULDCUT{
In practice, the candidates are generated at the Boogie IVL level, with respect to SMACK-generated memory regions and referencing the auxiliary variables introduced by our watchdog race-checking instrumentation.}

@@

\noindent\textbf{Limitations }
%
As a lockset analyzer, \whoop can be imprecise because a violation of the locking discipline does not always correspond to a real race (e.g.\ when lock-free synchronization is used). \whoop also uses over-approximation, which can be another source of imprecision. Furthermore, the tool does not check for dynamically created locks or for locks provided by external libraries, although the latter could be addressed by providing a mechanism for users to declare custom locks. We also do not currently treat interrupt handlers in a special way; we just assume that they can execute concurrently with any entry point. One way to address this is to model interrupt-specific kernel functions (e.g. for enabling/disabling interrupts).
